Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Cancer

### 2018-05-23 submission/manuscript.Rmd
add references to discussion

One limitation of ours and others research into the effect of SCFAs in colorectal cancer has been that all tumors are treated as the same type. However, there are known differences in the types of mutations that occur [] and treating all tumors as equal may actually hide any benefit that could be found in specific subsets of individuals. Similar to the idea of specific immunotherapy for specific tumors [], SCFAs may have beneficial effects for specific types of colorectal cancers. Future research will need to test if this is a valid hypothesis. Although there are limitations with the current research on SCFAs and colorectal tumors, evidence is building that these specific metabolites may not be protective. Yet, taxa that are associated with SCFA production are consistently higher in normal colons than those with carcinomas [].

The potential protection against colorectal cancers may not be from SCFAs even though taxa associated with their production are higher in individuals with normal colons versus patients with carcinomas []. Protection could be via a different pathway and other metabolites that have not been extensively studied. Alternatively, protection may not occur via a metabolite but instead occur through niche exclusion of mouth-associated microbes (e.g. *Fusobacterium*, *Porphyromonas*, *Parvimonas*, *Peptostreptococcus* []). The idea of niche exclusion is similar to how the community protects against *Clostridium difficile* infection with chronic inflammation replacing the role of antibiotics. Although we did not find lower concentrations of SCFAs associated with colorectal tumors, we think that there are many exciting new avenues to explore because of these observations.

>>>>

One limitation of ours and others research into the effect of SCFAs in colorectal cancer has been that all tumors are treated as the same type. However, there are known differences in the types of mutations that occur [@Fearon1990] and treating all tumors as equal may actually hide any benefit that could be found in specific subsets of individuals. Similar to the idea of specific immunotherapy for specific tumors [@Thomas2016], SCFAs may have beneficial effects for specific types of colorectal cancers. Future research will need to test if this is a valid hypothesis. Although there are limitations with the current research on SCFAs and colorectal tumors, evidence is building that these specific metabolites may not be protective. Yet, taxa that are associated with SCFA production are consistently higher in normal colons than those with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].

The potential protection against colorectal cancers may not be from SCFAs even though taxa associated with their production are higher in individuals with normal colons versus patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018]. Protection could be via a different pathway and other metabolites that have not been extensively studied. Alternatively, protection may not occur via a metabolite but instead occur through niche exclusion of mouth-associated microbes (e.g. *Fusobacterium*, *Porphyromonas*, *Parvimonas*, *Peptostreptococcus* [@Zeller2014; @Baxter2016]). The idea of niche exclusion is similar to how the community protects against *Clostridium difficile* infection with chronic inflammation replacing the role of antibiotics. Although we did not find lower concentrations of SCFAs associated with colorectal tumors, we think that there are many exciting new avenues to explore because of these observations.

### 2018-05-23 submission/manuscript.tex
add references to discussion

One limitation of ours and others research into the effect of SCFAs in
colorectal cancer has been that all tumors are treated as the same type.
However, there are known differences in the types of mutations that
occur {[}{]} and treating all tumors as equal may actually hide any
benefit that could be found in specific subsets of individuals. Similar
to the idea of specific immunotherapy for specific tumors {[}{]}, SCFAs
may have beneficial effects for specific types of colorectal cancers.
Future research will need to test if this is a valid hypothesis.
Although there are limitations with the current research on SCFAs and
colorectal tumors, evidence is building that these specific metabolites
may not be protective. Yet, taxa that are associated with SCFA
production are consistently higher in normal colons than those with
carcinomas {[}{]}.

The potential protection against colorectal cancers may not be from
SCFAs even though taxa associated with their production are higher in
individuals with normal colons versus patients with carcinomas {[}{]}.
Protection could be via a different pathway and other metabolites that
have not been extensively studied. Alternatively, protection may not
occur via a metabolite but instead occur through niche exclusion of
mouth-associated microbes (e.g. \emph{Fusobacterium},
\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus}
{[}{]}). The idea of niche exclusion is similar to how the community
protects against \emph{Clostridium difficile} infection with chronic
inflammation replacing the role of antibiotics. Although we did not find
lower concentrations of SCFAs associated with colorectal tumors, we
think that there are many exciting new avenues to explore because of
these observations.

@@

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (8, 9). In brief, this study used fecal
samples obtained at either a single cross-sectional time point (n=490)
or from before (pre-) and after (post-) treatment for their tumor
(n=67). For patients undergoing treatment for their tumor the length of
time between their initial and follow up sample ranged from 188 - 546
days. Our use of treatment has been previously defined as encompassing
removal of a tumor with or without chemotherapy and radiation (8).
Diagnosis of tumor was made by colonoscopic examination and
histopathological review of biopsies obtained (8, 9). The University of
Michigan Institutional Review Board approved the study and informed
consent was obtained from all participants in accordance to the
guidelines set out by the Helsinki Decleration.

\textbf{Measuring specific SCFAs.} Our protocol for the measurement of
acetate, butyrate, and propionate followed a previously published
protocol (10). The following changes to this protocol included the use
of frozen fecal samples suspended in 1ml of PBS instead of fecal
suspensions in DNA Genotek OmniGut tubes, and the use of the acutal
weight of fecal samples instead of the average weight for SCFA
concentration normalizations. These changes did not affect the overall
median concentrations of these SCFAs between the two studies (see Table
1 (10) and Figure 1 in this report).

\textbf{16s rRNA gene sequencing.} The workflow and processing have been
described previously (8, 11, 12). The major differences from these
previous reports include: the use of version 1.39.5 of the mothur
software package and clustering Operational Taxonomic Units (OTUs) at
97\% similarity used the OptClust algorithm (13).

\textbf{Generating imputed metagenomes.} The use of PICRUSt version
1.1.2 with the recommended standard operating protocol (14) was used.
Briefly, the mothur shared file and metadata was converted into a biom
formated table using the biom convert function, the subsequent biom file
was processed with the normalize\_by\_copy\_number.py function, and
subsequent imputed metagenomes created using the predict\_metagenomes.py
function.

\textbf{Obtaining OPFs from metagenomes.} A subset of the
cross-sectional group (n=490) containing a total of 85 individuals
(normal n=29 normal, adenoma n=28, and carcinoma n=28) was shotgun
sequenced on an Illumina HiSeq using 125 bp paired end reads and a
previously described method (15). Briefly, the sequences were quality
filtered and sequences aligning to the human genome were removed prior
to contig assembly with MEGAHIT (16). Open Reading Frames (ORFs) were
identified using Prodigal (17), counts generated using Diamond (18),
subsequent clustering into Operational Protein Families (OPFs) used
mmseq2 (19), and OPF alignment used the KEGG database (20).

\textbf{Pulling genes involved with SCFA synthesis.} Specific genes
located near the end of the pathways involved in the synthesis of
acetate, butyrate, and propionate were analyzed for any differences
between individuals with normal colons and those with tumors. These
genes were based on pathways from KEGG as well as previous research (20,
21) and a list can be found in the supplemental material {[}Table S1{]}.

\textbf{Random Forest Models.} The model was first trained on 80\% of
the data and then tested on the held out 20\% (80/20 split) using the
Random Forest algorithm for both classification and regression models
(22). This was repeated on 100 differen 80/20 splits of the data to
generate a reasonable range for the AUC of the model. The reported AUCs,
unless otherwise specified, are for the test sets. Classification models
were built to group normal versus adenoma and normal versus carcinoma or
high versus low SCFA concentrations.

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (23). Generally, differences between the
different disease groups used a Kruskal-Walis rank sum test with a
Dunn's post-hoc test. Where appropriate Benjamini-Hochberg was used to
correct for multiple comparisons (24). First, we assessed differences in
SCFA concentrations measured by HPLC between individuals with normal
colons and patients with tumors (adenoma or carcinoma). We then analyzed
whether SCFA concentrations changed in patients with an adenoma or
carcinoma pre- versus post-treatment. Next, we assessed whether OTUs
alone or OTUs and SCFAs were better able to classify individuals with
and without tumor using Random Forest models. Next, the imputed gene
counts of important mediators of SCFA creation were tested. Finally, the
counts generated for OPFs that matched important genes involved with
SCFA creation were analyzed. Finally, models to classify high or low
SCFA concentration based on 16S rRNA gene sequencing data were created
using the Random Forest algorithm.

@@

\hypertarget{ref-normalization_sze2017}{}
8. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
microbiota in patients after treatment for colonic lesions. Microbiome
\textbf{5}.
doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.

\hypertarget{ref-crc_model_baxter2016}{}
9. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\hypertarget{ref-scfa_measures_venkataraman2016}{}
10. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
\textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
Variable responses of human microbiomes to dietary supplementation with
resistant starch. Microbiome \textbf{4}.
doi:\href{https://doi.org/10.1186/s40168-016-0178-x}{10.1186/s40168-016-0178-x}.

\hypertarget{ref-Schloss2009}{}
11. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-source, platform-independent,
community-supported software for describing and comparing microbial
communities. Applied and Environmental Microbiology
\textbf{75}:7537--7541.
doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.

\hypertarget{ref-Kozich2013}{}
12. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
dual-index sequencing strategy and curation pipeline for analyzing
amplicon sequence data on the MiSeq illumina sequencing platform.
Applied and Environmental Microbiology \textbf{79}:5112--5120.
doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.

\hypertarget{ref-opticlust_westcott2017}{}
13. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
improved method for assigning amplicon-based sequence data to
operational taxonomic units. mSphere \textbf{2}:e00073--17.
doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.

\hypertarget{ref-Langille2013}{}
14. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
\textbf{McDonald D}, \textbf{Knights D}, \textbf{Reyes JA},
\textbf{Clemente JC}, \textbf{Burkepile DE}, \textbf{Thurber RLV},
\textbf{Knight R}, \textbf{Beiko RG}, \textbf{Huttenhower C}. 2013.
Predictive functional profiling of microbial communities using 16S rRNA
marker gene sequences. Nature Biotechnology \textbf{31}:814--821.
doi:\href{https://doi.org/10.1038/nbt.2676}{10.1038/nbt.2676}.

\hypertarget{ref-Hannigan2017}{}
15. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
\textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
\& the interactive dynamics of the colorectal cancer virome.
doi:\href{https://doi.org/10.1101/152868}{10.1101/152868}.

\hypertarget{ref-Li2015}{}
16. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
K}, \textbf{Lam T-W}. 2015. MEGAHIT: An ultra-fast single-node solution
for large and complex metagenomics assembly via succinct de bruijn
graph. Bioinformatics \textbf{31}:1674--1676.
doi:\href{https://doi.org/10.1093/bioinformatics/btv033}{10.1093/bioinformatics/btv033}.

\hypertarget{ref-Hyatt2010}{}
17. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
\textbf{Land ML}, \textbf{Larimer FW}, \textbf{Hauser LJ}. 2010.
Prodigal: Prokaryotic gene recognition and translation initiation site
identification. BMC Bioinformatics \textbf{11}:119.
doi:\href{https://doi.org/10.1186/1471-2105-11-119}{10.1186/1471-2105-11-119}.

\hypertarget{ref-Buchfink2014}{}
18. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
and sensitive protein alignment using DIAMOND. Nature Methods
\textbf{12}:59--60.
doi:\href{https://doi.org/10.1038/nmeth.3176}{10.1038/nmeth.3176}.

\hypertarget{ref-Steinegger2017}{}
19. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
sensitive protein sequence searching for the analysis of massive data
sets. Nature Biotechnology.
doi:\href{https://doi.org/10.1038/nbt.3988}{10.1038/nbt.3988}.

\hypertarget{ref-Kanehisa2015}{}
20. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
\textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
resource for gene and protein annotation. Nucleic Acids Research
\textbf{44}:D457--D462.
doi:\href{https://doi.org/10.1093/nar/gkv1070}{10.1093/nar/gkv1070}.

\hypertarget{ref-scfa_baxter2014}{}
21. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
\textbf{Schloss PD}. 2014. Structure of the gut microbiome following
colonization with human feces determines colonic tumor burden.
Microbiome \textbf{2}:20.
doi:\href{https://doi.org/10.1186/2049-2618-2-20}{10.1186/2049-2618-2-20}.

\hypertarget{ref-randomforest_citation_2002}{}
22. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
regression by randomForest. R News \textbf{2}:18--22.

\hypertarget{ref-r_citation_2017}{}
23. \textbf{R Core Team}. 2017. R: A language and environment for
statistical computing. R Foundation for Statistical Computing, Vienna,
Austria.

\hypertarget{ref-benjamini_controlling_1995}{}
24. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
false discovery rate: A practical and powerful approach to multiple
testing. Journal of the Royal Statistical Society Series B
(Methodological) \textbf{57}:289--300.

>>>>

One limitation of ours and others research into the effect of SCFAs in
colorectal cancer has been that all tumors are treated as the same type.
However, there are known differences in the types of mutations that
occur (8) and treating all tumors as equal may actually hide any benefit
that could be found in specific subsets of individuals. Similar to the
idea of specific immunotherapy for specific tumors (9), SCFAs may have
beneficial effects for specific types of colorectal cancers. Future
research will need to test if this is a valid hypothesis. Although there
are limitations with the current research on SCFAs and colorectal
tumors, evidence is building that these specific metabolites may not be
protective. Yet, taxa that are associated with SCFA production are
consistently higher in normal colons than those with carcinomas
(10--12).

The potential protection against colorectal cancers may not be from
SCFAs even though taxa associated with their production are higher in
individuals with normal colons versus patients with carcinomas (10--12).
Protection could be via a different pathway and other metabolites that
have not been extensively studied. Alternatively, protection may not
occur via a metabolite but instead occur through niche exclusion of
mouth-associated microbes (e.g. \emph{Fusobacterium},
\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus} (13,
14)). The idea of niche exclusion is similar to how the community
protects against \emph{Clostridium difficile} infection with chronic
inflammation replacing the role of antibiotics. Although we did not find
lower concentrations of SCFAs associated with colorectal tumors, we
think that there are many exciting new avenues to explore because of
these observations.

@@

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (15, 16). In brief, this study used fecal
samples obtained at either a single cross-sectional time point (n=490)
or from before (pre-) and after (post-) treatment for their tumor
(n=67). For patients undergoing treatment for their tumor the length of
time between their initial and follow up sample ranged from 188 - 546
days. Our use of treatment has been previously defined as encompassing
removal of a tumor with or without chemotherapy and radiation (15).
Diagnosis of tumor was made by colonoscopic examination and
histopathological review of biopsies obtained (15, 16). The University
of Michigan Institutional Review Board approved the study and informed
consent was obtained from all participants in accordance to the
guidelines set out by the Helsinki Decleration.

\textbf{Measuring specific SCFAs.} Our protocol for the measurement of
acetate, butyrate, and propionate followed a previously published
protocol (17). The following changes to this protocol included the use
of frozen fecal samples suspended in 1ml of PBS instead of fecal
suspensions in DNA Genotek OmniGut tubes, and the use of the acutal
weight of fecal samples instead of the average weight for SCFA
concentration normalizations. These changes did not affect the overall
median concentrations of these SCFAs between the two studies (see Table
1 (17) and Figure 1 in this report).

\textbf{16s rRNA gene sequencing.} The workflow and processing have been
described previously (15, 18, 19). The major differences from these
previous reports include: the use of version 1.39.5 of the mothur
software package and clustering Operational Taxonomic Units (OTUs) at
97\% similarity used the OptClust algorithm (20).

\textbf{Generating imputed metagenomes.} The use of PICRUSt version
1.1.2 with the recommended standard operating protocol (21) was used.
Briefly, the mothur shared file and metadata was converted into a biom
formated table using the biom convert function, the subsequent biom file
was processed with the normalize\_by\_copy\_number.py function, and
subsequent imputed metagenomes created using the predict\_metagenomes.py
function.

\textbf{Obtaining OPFs from metagenomes.} A subset of the
cross-sectional group (n=490) containing a total of 85 individuals
(normal n=29 normal, adenoma n=28, and carcinoma n=28) was shotgun
sequenced on an Illumina HiSeq using 125 bp paired end reads and a
previously described method (22). Briefly, the sequences were quality
filtered and sequences aligning to the human genome were removed prior
to contig assembly with MEGAHIT (23). Open Reading Frames (ORFs) were
identified using Prodigal (24), counts generated using Diamond (25),
subsequent clustering into Operational Protein Families (OPFs) used
mmseq2 (26), and OPF alignment used the KEGG database (27).

\textbf{Pulling genes involved with SCFA synthesis.} Specific genes
located near the end of the pathways involved in the synthesis of
acetate, butyrate, and propionate were analyzed for any differences
between individuals with normal colons and those with tumors. These
genes were based on pathways from KEGG as well as previous research (27,
28) and a list can be found in the supplemental material {[}Table S1{]}.

\textbf{Random Forest Models.} The model was first trained on 80\% of
the data and then tested on the held out 20\% (80/20 split) using the
Random Forest algorithm for both classification and regression models
(29). This was repeated on 100 differen 80/20 splits of the data to
generate a reasonable range for the AUC of the model. The reported AUCs,
unless otherwise specified, are for the test sets. Classification models
were built to group normal versus adenoma and normal versus carcinoma or
high versus low SCFA concentrations.

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (30). Generally, differences between the
different disease groups used a Kruskal-Walis rank sum test with a
Dunn's post-hoc test. Where appropriate Benjamini-Hochberg was used to
correct for multiple comparisons (31). First, we assessed differences in
SCFA concentrations measured by HPLC between individuals with normal
colons and patients with tumors (adenoma or carcinoma). We then analyzed
whether SCFA concentrations changed in patients with an adenoma or
carcinoma pre- versus post-treatment. Next, we assessed whether OTUs
alone or OTUs and SCFAs were better able to classify individuals with
and without tumor using Random Forest models. Next, the imputed gene
counts of important mediators of SCFA creation were tested. Finally, the
counts generated for OPFs that matched important genes involved with
SCFA creation were analyzed. Finally, models to classify high or low
SCFA concentration based on 16S rRNA gene sequencing data were created
using the Random Forest algorithm.

@@

\hypertarget{ref-Fearon1990}{}
8. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
colorectal tumorigenesis. Cell \textbf{61}:759--767.
doi:\href{https://doi.org/10.1016/0092-8674(90)90186-i}{10.1016/0092-8674(90)90186-i}.

\hypertarget{ref-Thomas2016}{}
9. \textbf{Thomas X}, \textbf{Heiblig M}. 2016. The development of
agents targeting the BCR-ABL tyrosine kinase as philadelphia
chromosome-positive acute lymphoblastic leukemia treatment. Expert
Opinion on Drug Discovery \textbf{11}:1061--1070.
doi:\href{https://doi.org/10.1080/17460441.2016.1227318}{10.1080/17460441.2016.1227318}.

\hypertarget{ref-Shah2017}{}
10. \textbf{Shah MS}, \textbf{DeSantis TZ}, \textbf{Weinmaier T},
\textbf{McMurdie PJ}, \textbf{Cope JL}, \textbf{Altrichter A},
\textbf{Yamal J-M}, \textbf{Hollister EB}. 2017. Leveraging
sequence-based faecal microbial community survey data to identify a
composite biomarker for colorectal cancer. Gut \textbf{67}:882--891.
doi:\href{https://doi.org/10.1136/gutjnl-2016-313189}{10.1136/gutjnl-2016-313189}.

\hypertarget{ref-normalization_Sze2017}{}
11. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
microbiota in patients after treatment for colonic lesions. Microbiome
\textbf{5}.
doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.

\hypertarget{ref-meta_analysis_crc_Sze2018}{}
12. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
rRNA gene surveys to identify reproducible biomarkers in individuals
with colorectal tumors.
doi:\href{https://doi.org/10.1101/285486}{10.1101/285486}.

\hypertarget{ref-Zeller2014}{}
13. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
\textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
\textbf{Amiot A}, \textbf{Bohm J}, \textbf{Brunetti F},
\textbf{Habermann N}, \textbf{Hercog R}, \textbf{Koch M},
\textbf{Luciani A}, \textbf{Mende DR}, \textbf{Schneider MA},
\textbf{Schrotz-King P}, \textbf{Tournigand C}, \textbf{Nhieu JTV},
\textbf{Yamada T}, \textbf{Zimmermann J}, \textbf{Benes V},
\textbf{Kloor M}, \textbf{Ulrich CM}, \textbf{Knebel Doeberitz M von},
\textbf{Sobhani I}, \textbf{Bork P}. 2014. Potential of fecal microbiota
for early-stage detection of colorectal cancer. Molecular Systems
Biology \textbf{10}:766--766.
doi:\href{https://doi.org/10.15252/msb.20145645}{10.15252/msb.20145645}.

\hypertarget{ref-Baxter2016}{}
14. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\hypertarget{ref-normalization_sze2017}{}
15. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
microbiota in patients after treatment for colonic lesions. Microbiome
\textbf{5}.
doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.

\hypertarget{ref-crc_model_baxter2016}{}
16. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\hypertarget{ref-scfa_measures_venkataraman2016}{}
17. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
\textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
Variable responses of human microbiomes to dietary supplementation with
resistant starch. Microbiome \textbf{4}.
doi:\href{https://doi.org/10.1186/s40168-016-0178-x}{10.1186/s40168-016-0178-x}.

\hypertarget{ref-Schloss2009}{}
18. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-source, platform-independent,
community-supported software for describing and comparing microbial
communities. Applied and Environmental Microbiology
\textbf{75}:7537--7541.
doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.

\hypertarget{ref-Kozich2013}{}
19. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
dual-index sequencing strategy and curation pipeline for analyzing
amplicon sequence data on the MiSeq illumina sequencing platform.
Applied and Environmental Microbiology \textbf{79}:5112--5120.
doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.

\hypertarget{ref-opticlust_westcott2017}{}
20. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
improved method for assigning amplicon-based sequence data to
operational taxonomic units. mSphere \textbf{2}:e00073--17.
doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.

\hypertarget{ref-Langille2013}{}
21. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
\textbf{McDonald D}, \textbf{Knights D}, \textbf{Reyes JA},
\textbf{Clemente JC}, \textbf{Burkepile DE}, \textbf{Thurber RLV},
\textbf{Knight R}, \textbf{Beiko RG}, \textbf{Huttenhower C}. 2013.
Predictive functional profiling of microbial communities using 16S rRNA
marker gene sequences. Nature Biotechnology \textbf{31}:814--821.
doi:\href{https://doi.org/10.1038/nbt.2676}{10.1038/nbt.2676}.

\hypertarget{ref-Hannigan2017}{}
22. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
\textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
\& the interactive dynamics of the colorectal cancer virome.
doi:\href{https://doi.org/10.1101/152868}{10.1101/152868}.

\hypertarget{ref-Li2015}{}
23. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
K}, \textbf{Lam T-W}. 2015. MEGAHIT: An ultra-fast single-node solution
for large and complex metagenomics assembly via succinct de bruijn
graph. Bioinformatics \textbf{31}:1674--1676.
doi:\href{https://doi.org/10.1093/bioinformatics/btv033}{10.1093/bioinformatics/btv033}.

\hypertarget{ref-Hyatt2010}{}
24. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
\textbf{Land ML}, \textbf{Larimer FW}, \textbf{Hauser LJ}. 2010.
Prodigal: Prokaryotic gene recognition and translation initiation site
identification. BMC Bioinformatics \textbf{11}:119.
doi:\href{https://doi.org/10.1186/1471-2105-11-119}{10.1186/1471-2105-11-119}.

\hypertarget{ref-Buchfink2014}{}
25. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
and sensitive protein alignment using DIAMOND. Nature Methods
\textbf{12}:59--60.
doi:\href{https://doi.org/10.1038/nmeth.3176}{10.1038/nmeth.3176}.

\hypertarget{ref-Steinegger2017}{}
26. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
sensitive protein sequence searching for the analysis of massive data
sets. Nature Biotechnology.
doi:\href{https://doi.org/10.1038/nbt.3988}{10.1038/nbt.3988}.

\hypertarget{ref-Kanehisa2015}{}
27. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
\textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
resource for gene and protein annotation. Nucleic Acids Research
\textbf{44}:D457--D462.
doi:\href{https://doi.org/10.1093/nar/gkv1070}{10.1093/nar/gkv1070}.

\hypertarget{ref-scfa_baxter2014}{}
28. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
\textbf{Schloss PD}. 2014. Structure of the gut microbiome following
colonization with human feces determines colonic tumor burden.
Microbiome \textbf{2}:20.
doi:\href{https://doi.org/10.1186/2049-2618-2-20}{10.1186/2049-2618-2-20}.

\hypertarget{ref-randomforest_citation_2002}{}
29. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
regression by randomForest. R News \textbf{2}:18--22.

\hypertarget{ref-r_citation_2017}{}
30. \textbf{R Core Team}. 2017. R: A language and environment for
statistical computing. R Foundation for Statistical Computing, Vienna,
Austria.

\hypertarget{ref-benjamini_controlling_1995}{}
31. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
false discovery rate: A practical and powerful approach to multiple
testing. Journal of the Royal Statistical Society Series B
(Methodological) \textbf{57}:289--300.
